---
title: "Extracting GDP per capita for countries using World Bank API"
output: html_notebook
---

# Aim 

The aim of this would be to extract the following as an exposure:

Exposures would be: 

* [GDP per capita](https://data.worldbank.org/indicator/NY.GDP.PCAP.PP.CD)   

There's an API/ R Web package described [here](https://cran.r-project.org/web/packages/wbstats/vignettes/Using_the_wbstats_package.html)

Let's try to use it to get this information 

```{r}
require(wbstats)
require(tidyverse)
require(readxl)

```


```{r}
gdp_percap_data <- wb(indicator= "NY.GDP.PCAP.PP.CD")

```

now let's see what we've got 

```{r}


gdp_percap_data %>% 
  ggplot(aes(x = date, y = value, group = country)) + 
  geom_line(alpha = 0.4) +
  scale_y_log10() +
  geom_line(aes(x = date, y = value), colour = "red", size = 1.5, data = gdp_percap_data %>% filter(iso2c == "GB")) 


```


Now density in 1990, 2000, 2010

```{r}
gdp_percap_data %>% 
  filter(date %in% c(1990, 2000, 2010)) %>% 
  ggplot(aes(x = value, group = date, fill = date)) + 
  geom_density(alpha = 0.4) + 
  scale_x_log10() + 
  labs(x = "GDP per capita", y = "proportion of countries")

```

# Other indicators 

I'm looking for variables that may explain gender inequalities in NCDs (and overall)


https://data.worldbank.org/indicator/NY.GDP.PCAP.PP.CD

```{r explore_wb_cachelist}
require(wbstats)
require(tidyverse)

str(wb_cachelist, max.level = 1)

gender_statistics_ids <- wb_cachelist$indicators[wb_cachelist$indicators$source == "Gender Statistics", "indicatorID"]

length(gender_statistics_ids)
```

So, there are 260 indicators listed as 'gender statistics'. The details of these are below


```{r get_gender_stats}
wb_cachelist$indicators %>% 
  filter(indicatorID %in% c(gender_statistics_ids))

```


# Note from Vittal 

See if the indicators from [this website](http://hdr.undp.org/en/composite/GII) are available. 

These are:

* Human Development Index
* Life expectancy at birth
* Expected years of schooling
* Mean years of schooling
* GNI per capita rank minus HDI rank
* HDI rank


# Identify coverage for selected variables 

I selected around 25 variables previously. These are indicated in an Excel workbook.

The aim of this research is to identify the temporal and geographic coverage of each of these variables 


First load the data 


```{r}
wb_indicator_selection <- readxl::read_excel("support/all_wb_indicators.xlsx")

indicators <- wb_indicator_selection %>% 
  filter(`Keep?` == TRUE) %>% 
  pull(indicatorID)


```

Now, for each of these indicators, want to extract 

```{r cache = TRUE}
selected_indicators <- wb(indicator = indicators) %>% as_tibble()

```


This request takes some time to run, so I'll save the result of the query elsewhere once produced


```{r}

write_rds(x = selected_indicators, path = "data/world_bank/extracted_indicators.RData")

```



```{r}
selected_indicators %>% 
  group_by(indicatorID, indicator) %>% 
  summarise(
    n_countries = length(unique(iso3c)), 
    n_years = length(unique(date)), 
    min_year = min(date), max_year = max(date)
  ) %>% 
  arrange(desc(n_years))

```

Want to identify countries by 3 digit code, and whether these are consistent with GBD 

We can start by looking at the country hierarchy info from GBD 


```{r}
gbd_locations <- readxl::read_excel("raw_data/IHME_GBD_2016_CODEBOOK/IHME_GBD_2016_GBD_LOCATION_HIERARCHY_Y2018M04D26.XLSX")

gbd_countries <- gbd_locations %>% 
  filter(location_id %in% 1:218) %>% 
  filter(level == 3) %>% 
  pull(location_name)
```


Now to identify if a match in GBD by country name only 


```{r}
selected_indicators %>% 
  group_by(country) %>% 
  summarise(min_date = min(date), max_date = max(date), n_years = length(unique(date))) %>% 
  ungroup() %>% 
  mutate(country_in_wb = country %in% gbd_countries) %>% 
  filter(country_in_wb)


```

